Depletion capabilities in the OpenMC Monte Carlo particle transport code

### 2020-05-19 manuscript.tex
Add reference for OpenMC 0.11
Work on methodology section

In the present work, we describe a new depletion solver in
OpenMC~\citep{romano2015ane1}, an open source, community-developed MC particle
transport code. To date, OpenMC has not had a built-in depletion solver. This
has led to numerous efforts in the community toward developing coupled
transport-depletion calculations with OpenMC~\citep{gul2017ane,
lanversin2017icone, lanversin2019phd, liu2019nst, zhuang2020pne, zhao2020ned,
zhao2020cpc, zhang2020ane}. While these efforts have proven the viability of
performing such coupled calculations with OpenMC, none of them has resulted in a
validated, openly-available, and generalized capability that OpenMC users can
take advantage of. The capability that is described herein is part of the OpenMC
project and can be found in version 0.11 of the code or later.

@@

The equation that governs the transmutation and decay of nuclides inside of an
irradiated environment can be written as
\begin{equation}
  \label{eq:depletion}
  \begin{split}
  \frac{dN_i(t)}{dt} = &\sum\limits_j \left [ f_{j \rightarrow i}
  \int_0^\infty dE \; \sigma_j (E, t) \phi(E,t) + \lambda_{j\rightarrow i}
  \right ] N_j(t) \\ &- \left [\int_0^\infty dE \; \sigma_i (E,t) \phi(E,t) +
  \sum\limits_j \lambda_{i\rightarrow j} \right ] N_i(t), \quad i=1,\dots,n
  \end{split}
\end{equation}
where
\begin{equation*}
  \begin{split}
  N_i(t) &\equiv \text{density of nuclide $i$ at time $t$} \\
  \sigma_i(E,t) &\equiv \text{transmutation cross section for nuclide $i$ at energy $E$ and time $t$} \\
  \phi(E,t) &\equiv \text{neutron flux at energy $E$ and time $t$} \\
  f_{j \rightarrow i} &\equiv \text{fraction of transmutation reactions in nuclide $j$ that produce nuclide $i$} \\
  \lambda_{j \rightarrow i} &\equiv \text{decay constant for decay modes in nuclide $j$ that produce nuclide $i$} \\
  n &\equiv \text{total number of nuclides.}
  \end{split}
\end{equation*}
Note that we have not included the spatial dependence of the flux or cross
sections. For \cref{eq:depletion} to be accurate, the spatial variation of the
reaction rates over the region which is being depleted should be small. For a
reactor model, this may entail using a unique material in each fuel pin that is
depleted separately. Ultimately, the user is responsible for deciding what level
of discretization is necessary for the problem at hand.

\Cref{eq:depletion} simply states that the rate of change of $N_i$ is equal to
the production rate minus the loss rate. Because the equation for $N_i$ depends
on the density other nuclides, $\{N_j \,|\, j \ne i\}$, \cref{eq:depletion}
represents a system of first-order differential equations. To form a proper
initial value problem, we also need the nuclide densities at time 0:
\begin{equation}
    N_i(0) = N_{i,0}.
\end{equation}
These equations can be written more compactly in matrix notation as
\begin{equation}
  \label{eq:depletion-matrix}
  \frac{d\mathbf{n}}{dt} = \mathbf{A}(\mathbf{n},t)\mathbf{n}, \quad \mathbf{n}(0) =
  \mathbf{n}_0
\end{equation}
where
\begin{equation}
  \mathbf{n} = \begin{pmatrix} N_1 \\ N_2 \\ \vdots \\ N_n \end{pmatrix}, \quad
  \mathbf{n}_0 = \begin{pmatrix} N_{1,0} \\ N_{2,0} \\ \vdots \\ N_{n,0} \end{pmatrix}
\end{equation}
and $\mathbf{A}(\mathbf{n},t) \in \mathbb{R}^{n\times n}$ is the burnup matrix
containing the decay and transmutation coefficients. Note that the burnup matrix
depends on $\mathbf{n}$ because the solution to the transport equation depends
on the nuclide densities.

\subsection{Time integration}
\label{sec:time_integration}

Mention Isotalo papers

A variety of numerical methods exist for solving \cref{eq:depletion-matrix}. The
simplest such method, known as the "predictor" method, is to divide the overall
time interval of interest $[0,t]$ into smaller timesteps over which it is
assumed that the burnup matrix is constant. Let $t \in [t_i, t_i + h]$ be one
such timestep. Over the timestep, the solution to \cref{eq:depletion-matrix} can
be written analytically using the matrix exponential
\begin{equation}
    \mathbf{A}_i = \mathbf{A}(\mathbf{n}_i, t_i) \\
    \mathbf{n}_{i+1} = e^{\mathbf{A}_i h} \mathbf{n}_i
\end{equation}
where $\mathbf{n}_i \equiv \mathbf{n}(t_i)$. The exponential of a matrix
$\mathbf{X}$ is defined by the power series expansion
\begin{equation}
    e^{\mathbf{X}} = \sum\limits_{k=0}^\infty \frac{1}{k!} \left ( \mathbf{X}
    \right )^k
\end{equation}
where $\mathbf{X}^0 = \mathbf{I}$. A series of so-called predictor-corrector
methods that use multiple stages offer improved accuracy over the predictor
method. The simplest of these methods, the CE/CM algorithm, is defined as
\begin{equation}
    \mathbf{n}_{i+1/2} = e^{\frac{h}{2}\mathbf{A}(\mathbf{n}_i, t_i)} \mathbf{n}_i \\
    \mathbf{n}_{i+1} = e^{h \mathbf{A}(\mathbf{n}_{i+1/2},t_{i+1/2})} \mathbf{n}_i
\end{equation}
Here, the value of $\mathbf{n}$ at the midpoint is estimated using
$\mathbf{A}$ evaluated at the beginning of the timestep. Then,
$\mathbf{A}$ is evaluated using the densities at the midpoint and used to
integrate over the entire timestep.

@@

As we saw in \cref{sec:time_integration}, numerically integrating
\cref{eq:depletion-matrix} requires evaluating one or more matrix exponentials.
OpenMC uses the Chebyshev rational approximation method (CRAM), which was
introduced in a series of papers by Pusa~\citep{pusa2010nse,pusa2011nse}, to
evaluate matrix exponentials. In particular, OpenMC utilizes an incomplete
partial fraction (IPF) form~\citep{pusa2016nse} of CRAM that provides a good
balance of numerical stability and efficiency. In this representation the matrix
exponential is approximated as
\begin{equation}
    e^{\mathbf{A}t} \approx \alpha_0 \prod\limits_{\ell=1}^{k/2} \left (
    \mathbf{I} + 2 \text{Re} \left ( \widetilde{\alpha}_\ell \left (\mathbf{A}t
    - \theta_\ell \mathbf{I} \right )^{-1} \right ) \right )
\end{equation}
where $k$ is the order of the approximation and $\alpha_0$,
$\widetilde{\alpha}_\ell$, and $\theta_\ell$ are coefficients that have been
tabulated for orders up to $k=48$. Rather than computing the full approximation
and then multiplying it by a vector, the following algorithm\footnote{The
original description of the algorithm presented by \citet{pusa2016nse} contains
a typo.} is used to incrementally apply the terms within the product:
\begin{algorithm}[H]
  \caption{Incomplete partial fraction form of CRAM.}
  \label{alg:cram}
  \begin{algorithmic}[1]
    \State $\mathbf{n} \gets \mathbf{n_0}$
    \For{$\ell \gets 1$ \textbf{to} $k/2$}
      \State $\mathbf{n} \gets \mathbf{n} + 2\text{Re}(\widetilde{\alpha}_\ell
        (\mathbf{A}t - \theta_\ell)^{-1})\mathbf{n}$
    \EndFor
    \State $\mathbf{n} \gets \alpha_0 \mathbf{n}$
  \end{algorithmic}
\end{algorithm}
The $k$\ th order approximation for CRAM requires solving $k/2$ sparse linear
systems. OpenMC relies on functionality from scipy.sparse.linalg for solving the
linear systems.

@@

>>>>

In the present work, we describe a new depletion solver in
OpenMC~\citep{romano2015ane1}, an open source, community-developed MC particle
transport code. To date, OpenMC has not had a built-in depletion solver. This
has led to numerous efforts in the community toward developing coupled
transport-depletion calculations with OpenMC~\citep{gul2017ane,
lanversin2017icone, lanversin2019phd, liu2019nst, zhuang2020pne, zhao2020ned,
zhao2020cpc, zhang2020ane}. While these efforts have proven the viability of
performing such coupled calculations with OpenMC, none of them has resulted in a
validated, openly-available, and generalized capability that OpenMC users can
take advantage of. The capability that is described herein is part of the OpenMC
project and can be found in version 0.11~\citep{openmc-0110} of the code or
later.

@@

The equation that governs the transmutation and decay of nuclides inside of an
irradiated environment can be written as
\begin{equation}
  \label{eq:depletion}
  \begin{split}
  \frac{dN_i(t)}{dt} = &\sum\limits_j \left [ f_{j \rightarrow i}
  \int_0^\infty dE \; \sigma_j (E, t) \phi(E,t) + \lambda_{j\rightarrow i}
  \right ] N_j(t) \\ &- \left [\int_0^\infty dE \; \sigma_i (E,t) \phi(E,t) +
  \sum\limits_j \lambda_{i\rightarrow j} \right ] N_i(t), \quad i=1,\dots,n
  \end{split}
\end{equation}
where
\begin{equation*}
  \begin{split}
  N_i(t) &\equiv \text{density of nuclide $i$ at time $t$} \\
  \sigma_i(E,t) &\equiv \text{transmutation cross section for nuclide $i$ at energy $E$ and time $t$} \\
  \phi(E,t) &\equiv \text{neutron flux at energy $E$ and time $t$} \\
  f_{j \rightarrow i} &\equiv \text{fraction of transmutation reactions in nuclide $j$ that produce nuclide $i$} \\
  \lambda_{j \rightarrow i} &\equiv \text{decay constant for decay modes in nuclide $j$ that produce nuclide $i$} \\
  n &\equiv \text{total number of nuclides.}
  \end{split}
\end{equation*}
Note that we have not included the spatial dependence of the flux or cross
sections. For \cref{eq:depletion} to be accurate, the spatial variation of the
reaction rates over the region which is being depleted should be small. For a
reactor model, this may entail using a unique material in each fuel pin that is
depleted separately. Ultimately, the user is responsible for deciding what level
of discretization is adequate for the problem at hand.

\Cref{eq:depletion} simply states that the rate of change of $N_i$ is equal to
the production rate minus the loss rate. Because the equation for $N_i$ depends
on the density of other nuclides, $\{N_j \,|\, j \ne i\}$, \cref{eq:depletion}
represents a system of first-order differential equations. To form a proper
initial value problem, the nuclide densities at time 0 must be specified:
\begin{equation}
    N_i(0) = N_{i,0}.
\end{equation}
These equations can be written more compactly in matrix notation as
\begin{equation}
  \label{eq:depletion-matrix-t}
  \frac{d\mathbf{n}}{dt} = \mathbf{A}(\mathbf{n},t)\mathbf{n}, \quad \mathbf{n}(0) =
  \mathbf{n}_0
\end{equation}
where
\begin{equation}
  \mathbf{n} = \begin{pmatrix} N_1 \\ N_2 \\ \vdots \\ N_n \end{pmatrix}, \quad
  \mathbf{n}_0 = \begin{pmatrix} N_{1,0} \\ N_{2,0} \\ \vdots \\ N_{n,0} \end{pmatrix}
\end{equation}
and $\mathbf{A}(\mathbf{n},t) \in \mathbb{R}^{n\times n}$ is the burnup matrix
containing the decay and transmutation coefficients. Note that the burnup matrix
depends on $\mathbf{n}$ because the solution to the transport equation depends
on the nuclide densities, thereby making \cref{eq:depletion-matrix-t} nonlinear.

Although the neutron transport equation is time-dependent, the timescale over
which material compositions change is sufficiently long such that the transport
equation can be solved as a time-independent equation. Without loss of
generality, let us write the burnup matrix solely as a function of $\mathbf{n}$
so that \cref{eq:depletion-matrix-t} becomes
\begin{equation}
  \label{eq:depletion-matrix}
  \frac{d\mathbf{n}}{dt} = \mathbf{A}(\mathbf{n})\mathbf{n}, \quad \mathbf{n}(0) =
  \mathbf{n}_0.
\end{equation}
This better reflects the fact that the time, $t$, is generally not an input to
the transport equation, and hence the burnup matrix is uniquely determined by
the value of $\mathbf{n}$.

The principle difficulty of solving \cref{eq:depletion-matrix} is the fact that
the elements of $\mathbf{A}$ are not constant with respect to time. If
$\mathbf{A}$ is constant, \cref{eq:depletion-matrix} admits a closed-form
solution,
\begin{equation}
  \mathbf{n}(t) = \exp \left (\mathbf{A} t \right ) \mathbf{n}_0.
\end{equation}
The exponential of a matrix $\mathbf{X}$ is defined by the power series
expansion
\begin{equation}
    \exp(\mathbf{X}) = \sum\limits_{k=0}^\infty \frac{1}{k!} \mathbf{X}^k
\end{equation}
where $\mathbf{X}^0 = \mathbf{I}$. Most numerical methods for solving
\cref{eq:depletion-matrix} involve two separate components:
\begin{enumerate}
  \item Approximating \cref{eq:depletion-matrix} in a form that involves taking
  one or more matrix exponentials, and
  \item Evaluating the matrix exponential or, equivalently, the action of a
  matrix exponential on a vector.
\end{enumerate}
We will henceforth refer to the first component as \emph{time integration}. Let
us now consider how each of these components are treated in OpenMC.

\subsection{Time Integration}
\label{sec:time_integration}

A variety of numerical methods exist for solving \cref{eq:depletion-matrix}. The
simplest such method, known as the "predictor" method, is to divide the overall
time interval of interest $[0,t]$ into smaller timesteps over which it is
assumed that the burnup matrix is constant. Let $t \in [t_i, t_i + h]$ be one
such timestep. Over the timestep, the solution to \cref{eq:depletion-matrix} can
be written analytically using the matrix exponential
\begin{align}
    \mathbf{A}_i &= \mathbf{A}(\mathbf{n}_i) \\
    \mathbf{n}_{i+1} &= \exp(\mathbf{A}_i h) \mathbf{n}_i
\end{align}
where $\mathbf{n}_i \equiv \mathbf{n}(t_i)$. A series of so-called predictor-corrector
methods that use multiple stages offer improved accuracy over the predictor
method. The simplest of these methods, the CE/CM algorithm, is defined as
\begin{align}
    \mathbf{n}_{i+1/2} &= \exp \left (\frac{h}{2}\mathbf{A}(\mathbf{n}_i) \right) \mathbf{n}_i \\
    \mathbf{n}_{i+1} &= \exp \left(h \mathbf{A}(\mathbf{n}_{i+1/2}) \right) \mathbf{n}_i
\end{align}
Here, the value of $\mathbf{n}$ at the midpoint is estimated using
$\mathbf{A}$ evaluated at the beginning of the timestep. Then,
$\mathbf{A}$ is evaluated using the densities at the midpoint and used to
integrate over the entire timestep.

@@

As we saw in \cref{sec:time_integration}, numerically integrating
\cref{eq:depletion-matrix} requires evaluating one or more matrix exponentials.
OpenMC uses the Chebyshev rational approximation method (CRAM), which was
introduced in a series of papers by Pusa~\citep{pusa2010nse,pusa2011nse}, to
evaluate matrix exponentials. In particular, OpenMC utilizes an incomplete
partial fraction (IPF) form~\citep{pusa2016nse} of CRAM that provides a good
balance of numerical stability and efficiency. In this representation the matrix
exponential is approximated as
\begin{equation}
    \exp(\mathbf{A}t) \approx \alpha_0 \prod\limits_{\ell=1}^{k/2} \left (
    \mathbf{I} + 2 \text{Re} \left ( \widetilde{\alpha}_\ell \left (\mathbf{A}t
    - \theta_\ell \mathbf{I} \right )^{-1} \right ) \right )
\end{equation}
where $k$ is the order of the approximation and $\alpha_0$,
$\widetilde{\alpha}_\ell$, and $\theta_\ell$ are coefficients that have been
tabulated for orders up to $k=48$. Rather than computing the full approximation
and then multiplying it by a vector, the following algorithm\footnote{The
original description of the algorithm presented by \citet{pusa2016nse} contains
a typo.} is used to incrementally apply the terms within the product:
\begin{algorithm}[H]
  \caption{Incomplete partial fraction form of CRAM.}
  \label{alg:cram}
  \begin{algorithmic}[1]
    \State $\mathbf{n} \gets \mathbf{n_0}$
    \For{$\ell \gets 1$ \textbf{to} $k/2$}
      \State $\mathbf{n} \gets \mathbf{n} + 2\text{Re}(\widetilde{\alpha}_\ell
        (\mathbf{A}t - \theta_\ell)^{-1})\mathbf{n}$
    \EndFor
    \State $\mathbf{n} \gets \alpha_0 \mathbf{n}$
  \end{algorithmic}
\end{algorithm}
The $k$th order approximation for CRAM requires solving $k/2$ sparse linear
systems. OpenMC relies on functionality from scipy.sparse.linalg for solving the
linear systems.

\subsection{Implementation}

In-memory handling through Python API. This allows time loading nuclear data to
be amortized since it only needs to be done once.

@@

\subsection{Data Preparation}

Using ENDF/B-VII.1 files from MCNP distribution. However, for light water
$S(\alpha,\beta)$, Serpent is not capable of handling continuous format, so the
file from ENDF/B-VII.0 was used. The evaluation didn't change between VII.0 and
VII.1, but the processing was different. For Serpent, decay and neutron fission
yield sublibraries are concatenated.

\subsection{}
